---
title: "Traffic Accident Analysis in Addis Ababa"
author: "Nasir Hussain"
date: "2024-11-07"
output: html_document
---
# Introduction

In this report, we analyze traffic accident data collected from the Addis Ababa Sub-city police departments. The dataset spans from 2017 to 2020 and includes various features such as driver demographics, vehicle information, accident conditions, and accident severity. This analysis aims to uncover patterns and key factors influencing traffic accident severity in Addis Ababa.


## Objective

This project seeks to answer the following research questions:

1. What are the primary causes of road traffic accidents in Addis Ababa?

2. How do environmental and temporal factors (e.g., weather, time of day) influence accident severity?

3. What patterns can we observe based on driver demographics and vehicle information?

4. Is there a correlation between vehicle-related factors (e.g., type of vehicle, service year) and accident severity?

5. Are there time-based trends in accident occurrence, such as peaks during certain times of the day or days of the week?

6. How do different road features (e.g., road alignment, type of junction) affect accident frequency and severity?

7. Is there a relationship between the number of vehicles involved in an accident and the severity of the outcome?

8. How does pedestrian movement influence accident outcomes?

9. Does the driving experience of the driver affect the likelihood of severe accidents?

10. Can we predict accident severity based on the driver’s and vehicle’s features?



# Data Wrangling and Cleaning
```{r}
# Load necessary libraries
library(tidyverse)
library(lubridate)
library(ggplot2)
library(dplyr)

```

```{r}


# Load the dataset
accidents <- read.csv("RTA Dataset.csv")

# Inspect the dataset
str(accidents)
summary(accidents)

```
```{r}
# Ensure the columns has correct data types

# Convert columns with categorical data to factors
accidents$Day_of_week <- factor(accidents$Day_of_week)
accidents$Sex_of_driver <- factor(accidents$Sex_of_driver)
accidents$Age_band_of_driver <- factor(accidents$Age_band_of_driver)
accidents$Educational_level <- factor(accidents$Educational_level)
accidents$Driving_experience <- factor(accidents$Driving_experience)
accidents$Type_of_vehicle <- factor(accidents$Type_of_vehicle)
accidents$Owner_of_vehicle <- factor(accidents$Owner_of_vehicle)
accidents$Service_year_of_vehicle <- factor(accidents$Service_year_of_vehicle)
accidents$Defect_of_vehicle <- factor(accidents$Defect_of_vehicle)
accidents$Area_accident_occured <- factor(accidents$Area_accident_occured)
accidents$Lanes_or_Medians <- factor(accidents$Lanes_or_Medians)
accidents$Road_allignment <- factor(accidents$Road_allignment)
accidents$Types_of_Junction <- factor(accidents$Types_of_Junction)
accidents$Road_surface_type <- factor(accidents$Road_surface_type)
accidents$Road_surface_conditions <- factor(accidents$Road_surface_conditions)
accidents$Light_conditions <- factor(accidents$Light_conditions)
accidents$Weather_conditions <- factor(accidents$Weather_conditions)
accidents$Type_of_collision <- factor(accidents$Type_of_collision)
accidents$Vehicle_movement <- factor(accidents$Vehicle_movement)
accidents$Casualty_class <- factor(accidents$Casualty_class)
accidents$Sex_of_casualty <- factor(accidents$Sex_of_casualty)
accidents$Age_band_of_casualty <- factor(accidents$Age_band_of_casualty)
accidents$Casualty_severity <- factor(accidents$Casualty_severity)
accidents$Work_of_casuality <- factor(accidents$Work_of_casuality)
accidents$Fitness_of_casuality <- factor(accidents$Fitness_of_casuality)
accidents$Pedestrian_movement <- factor(accidents$Pedestrian_movement)
accidents$Cause_of_accident <- factor(accidents$Cause_of_accident)
accidents$Accident_severity <- factor(accidents$Accident_severity)

# Convert Time to POSIXct (time format)
accidents$Time <- as.POSIXct(accidents$Time, format="%H:%M:%S")


```
```{r}
#handling missing values

# Drop rows with missing values
accidents_cleaned <- na.omit(accidents)

# Inspect the cleaned dataset
str(accidents_cleaned)
summary(accidents_cleaned)


```
# 1. What factors contribute most to road traffic accidents in Addis Ababa?

```{r}
# Plot the distribution of accidents by driver age group
ggplot(accidents_cleaned, aes(x = Age_band_of_driver)) +
  geom_bar(fill = "skyblue", color = "black") +
  labs(title = "Accidents by Age Band of Driver", 
       x = "Age Band of Driver", y = "Count of Accidents")

# Plot accidents by sex of the driver
ggplot(accidents_cleaned, aes(x = Sex_of_driver)) +
  geom_bar(fill = "orange", color = "black") +
  labs(title = "Accidents by Sex of Driver", 
       x = "Sex of Driver", y = "Count of Accidents")

# Plot accidents by driving experience
ggplot(accidents_cleaned, aes(x = Driving_experience)) +
  geom_bar(fill = "purple", color = "black") +
  labs(title = "Accidents by Driving Experience", 
       x = "Driving Experience", y = "Count of Accidents")

# Plot accidents by type of vehicle
ggplot(accidents_cleaned, aes(x = Type_of_vehicle)) +
  geom_bar(fill = "lightgreen", color = "black") +
  labs(title = "Accidents by Type of Vehicle", 
       x = "Type of Vehicle", y = "Count of Accidents")+
    theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis labels by 45 degrees


```
# 2. How do accidents vary by day of the week?

```{r}
# Plot accidents by day of the week
ggplot(accidents_cleaned, aes(x = Day_of_week)) +
  geom_bar(fill = "salmon", color = "black") +
  labs(title = "Accidents by Day of the Week", x = "Day of the Week", y = "Count of Accidents")


```
# 3. Is there any significant relationship between the driver’s age and the severity of accidents?
```{r}
# Plot age band vs. accident severity
ggplot(accidents_cleaned, aes(x = Age_band_of_driver, fill = Accident_severity)) +
  geom_bar(position = "dodge", color = "black") +
  labs(title = "Accident Severity by Age Band of Driver", 
       x = "Age Band of Driver", y = "Count of Accidents")

```
# 4.Does the type of vehicle affect the number of casualties in accidents?

```{r}
# Plot type of vehicle vs. number of casualties
ggplot(accidents_cleaned, aes(x = Type_of_vehicle, y = Number_of_casualties)) +
  geom_boxplot(fill = "lightgreen", color = "black") +
  labs(title = "Casualties by Type of Vehicle", 
       x = "Type of Vehicle", y = "Number of Casualties")

```

```{r message=FALSE, warning=FALSE}
# 1. Accidents by Day of the Week
day_of_week_counts <- table(accidents_cleaned$Day_of_week)  # Count the occurrences

# Print the actual counts
print(day_of_week_counts)

# Plot the accidents by Day of the Week
ggplot(accidents_cleaned, aes(x = Day_of_week)) +
  geom_bar(fill = "salmon", color = "black") +
  geom_text(stat = "count", aes(label = ..count..), vjust = -0.5) +
  labs(title = "Accidents by Day of the Week", x = "Day of the Week", y = "Count of Accidents")

```

