---
title: "Traffic Accident Analysis in Addis Ababa"
date: "2024-11-07"
output:
  html_document: default
  pdf_document: default
---
# Background and Problem definitions

Traffic accidents are among the leading causes of injury and death globally. Most importantly, the study of causes and antecedents of these mishaps will go a long way in enhancing road safety, and hence decreasing the fatality rate. In this research, the Road Traffic Accident Dataset of Addis Ababa City is employed to analyze the correlation between antecedent conditions such as weather conditions, driver’s experience, and the Road Alignment and Traffic congestion, with the extent of accident severity.

Specific inputs from the data set include details of over 12000 accidents that occurred between 2017 and 2020, and the circumstances when they occurred including time, location, type of vehicle and conditions at the time of the accident. Particularly, this work will be aimed at identifying specific characteristics of severe accident types because they have the highest impact on public health.

The issue under focus in this study is the analysis of severe road traffic accidents to determine their main antecedents. Thus, working with this dataset, the aim is to identify potential patterns as to how different variables such as weather condition, drivers’ experience, road gradients and much more are linked to the severity of the accidents. The findings generated from the analysis may serve to enhance the creation of safer prevention strategies that may assist policymakers, traffic authorities, and the public to minimize the number of fatal accidents on the road.

## Objective

This project seeks to answer the following research questions:
1. What are the most common causes of road traffic accidents in Addis Ababa, and how frequently do they occur?

2. How the Accidents are distributed among the male and female Drivers?

3. How the Accidents are distributed among the Age Band of Drivers?

4. What is the realtion between the experince of driver and the number of accidents?

5. What is the distribution of road accidents by the type of vehicle involved?

6. Is there a correlation between road surface conditions and accident severity?

7. Are there any noticeable trends in accident severity by road alignment type (e.g., curved, flat terrain)?

8. What factors most predict severe accidents (e.g., weather, driving experience, road alignment)?



# Data Wrangling and Cleaning
```{r}
# Load necessary libraries
library(tidyverse)
library(lubridate)
library(ggplot2)
library(dplyr)

```

```{r}


# Load the dataset
accidents <- read.csv("RTA Dataset.csv")

# Inspect the dataset
str(accidents)
summary(accidents)

```
The columns are being dropped because they either provide irrelevant information for analyzing the main factors of road traffic accidents or contain data that is not central to understanding accident causes, severity, or patterns. This simplification helps focus on the most impactful variables for the analysis.


```{r}
# Drop the specified columns
cols_to_drop <- c("Time", "Educational_level", "Owner_of_vehicle", "Lanes_or_Medians", 
                  "Types_of_Junction", "Casualty_class", "Sex_of_casualty", 
                  "Age_band_of_casualty", "Casualty_severity", "Work_of_casuality", 
                  "Fitness_of_casuality", "Pedestrian_movement")

# Remove the columns from the dataset
new_accidents <- accidents[, !(names(accidents) %in% cols_to_drop)]

# Check the remaining columns
remaining_columns <- colnames(new_accidents)
print("Remaining Columns After Dropping:")
print(remaining_columns)
```
## Handling missing values
```{r}
# Drop rows with missing values
accidents_clean <- na.omit(new_accidents)

# Print the first few rows to check
head(accidents_clean)

```
## 1.What are the most common causes of road traffic accidents in Addis Ababa, and how frequently do they occur?

```{r}
# Count the total number of accidents for each cause
cause_accidents_count <- table(accidents_clean$Cause_of_accident)

# Convert the table into a data frame for easier manipulation
cause_accidents_df <- as.data.frame(cause_accidents_count)
colnames(cause_accidents_df) <- c("Cause_of_accident", "Total_Accidents")

# Display the result
print(cause_accidents_df)

# Visualize the count of accidents by cause
library(ggplot2)
ggplot(cause_accidents_df, aes(x = reorder(Cause_of_accident, -Total_Accidents), y = Total_Accidents)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  labs(title = "Total Number of Accidents by Cause", 
       x = "Cause of Accident", 
       y = "Total Number of Accidents") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) # Rotate x labels for better readability

```

## Answer

The most frequent causes of accidents in Addis Ababa are linked to behaviors like changing lanes improperly, not maintaining adequate distance, and careless driving. Among these, No Distancing is the most prevalent, occurring in over 2,200 incidents. Other common causes include lane changes, careless driving, and lack of priority to vehicles and pedestrians.



## 2. How the Accidents are distributed among the male and female Drivers?

```{r}
sex_counts <- table(accidents_clean$Sex_of_driver)
ggplot(as.data.frame(sex_counts), aes(x = "", y = Freq, fill = Var1)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar("y") +
  labs(title = "Accidents by Sex of Driver") +
  theme_void()
print(sex_counts)

```

## Answer

The majority of road traffic accidents in Addis Ababa involve male drivers, accounting for a significantly higher number of accidents compared to female drivers. The unknown category also represents a small portion of the total accidents


## 3.How the Accidents are distributed among the Age Band of Drivers?


```{r}
ggplot(accidents_clean, aes(x = Age_band_of_driver)) +
  geom_bar(fill = "orange") +
  labs(title = "Total Number of Accidents by Age Band of Driver", x = "Age Band of Driver", y = "Total Number of Accidents") +
  theme_minimal()

# Print total accidents by age band
table(accidents_clean$Age_band_of_driver)




```

## Answer

The data shows a higher number of accidents among younger drivers (18-30 years) and those in the 31-50 years group. This suggests that driving experience (or lack of it) is a significant factor in accident occurrence, with younger drivers facing higher risks



## 4. What is the realtion between the experince of driver and the number of accidents?


```{r}
ggplot(accidents_clean, aes(x = Driving_experience)) +
  geom_bar(fill = "lightgreen") +
  labs(title = "Total Number of Accidents by Driving Experience", x = "Driving Experience", y = "Total Number of Accidents") +
  theme_minimal()

# Print total accidents by driving experience
table(accidents_clean$Driving_experience)




```

## Answer
The data shows that both inexperienced drivers (below 1 year) and very experienced drivers (above 10 years) are involved in the highest number of accidents. This suggests that driving experience is a key factor in accident occurrence, with both ends of the spectrum (new and highly experienced drivers) potentially facing higher risks. The middle range (2-5 years and 5-10 years) shows a moderate number of accidents, suggesting these drivers are somewhat more experienced and likely make fewer errors on the road.






## 5. What is the distribution of road accidents by the type of vehicle involved?


```{r}
# Load necessary libraries
library(ggplot2)
library(dplyr)

# Group by 'Type_of_vehicle' and count the number of accidents
vehicle_accidents <- accidents_clean %>%
  group_by(Type_of_vehicle) %>%
  summarise(accidents_count = n())

# View the result
print(vehicle_accidents)

# Create a bar plot to visualize the distribution
ggplot(vehicle_accidents, aes(x = Type_of_vehicle, y = accidents_count, fill = Type_of_vehicle)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(title = "Distribution of Road Accidents by Vehicle Type",
       x = "Type of Vehicle", y = "Number of Accidents") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

## Answer
Automobiles and Lorries (especially larger ones) dominate the number of accidents in Addis Ababa.
Public transport vehicles also contribute notably to accident statistics.
Motorcycles and smaller vehicles like Bajajs have a smaller presence but still contribute to the overall accident count.



## 6.Is there a correlation between road surface conditions and accident severity?

```{r}
# Create a contingency table for Road Surface Conditions vs Accident Severity
contingency_table <- table(accidents_clean$Road_surface_conditions, accidents_clean$Accident_severity)

# View the contingency table
print(contingency_table)

# Perform a Chi-squared test
chi_square_test <- chisq.test(contingency_table)

# Output the Chi-squared test result
print(chi_square_test)

# Visualization - Bar plot
ggplot(accidents_clean, aes(x = Road_surface_conditions, fill = Accident_severity)) +
  geom_bar(position = "fill") +
  theme_minimal() +
  labs(title = "Correlation Between Road Surface Conditions and Accident Severity",
       x = "Road Surface Conditions", y = "Proportion of Accident Severity") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

## Answer
The Graph above shows that accidents on Roads that are Dry or Wer/damp casues serious injury as comapred to those of others.


## 7.Are there any noticeable trends in accident severity by road alignment type (e.g., curved, flat terrain)?

```{r}
# Create a contingency table for Road Alignment vs Accident Severity
contingency_table_alignment <- table(accidents_clean$Road_allignment, accidents_clean$Accident_severity)

# View the contingency table
print(contingency_table_alignment)

# Perform a Chi-squared test
chi_square_test_alignment <- chisq.test(contingency_table_alignment)

# Output the Chi-squared test result
print(chi_square_test_alignment)

# Visualization - Bar plot
ggplot(accidents_clean, aes(x = Road_allignment, fill = Accident_severity)) +
  geom_bar(position = "fill") +
  theme_minimal() +
  labs(title = "Accident Severity by Road Alignment Type",
       x = "Road Alignment Type", y = "Proportion of Accident Severity") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

## Answer
As shown in the graph "Steep grade upward with mountainous terrain" have most of serious injuries in accidents. followed by "Tangent road wih rolling terrain" and "sharp reverse curve".

## 8.What factors most predict severe accidents (e.g., weather, driving experience, road alignment)?

```{r}
# Convert Accident Severity into a binary outcome (Severe vs. Non-severe)
accidents_clean$Severity_binary <- ifelse(accidents_clean$Accident_severity %in% c("Serious Injury", "Fatal"), 1, 0)

# Build the logistic regression model
log_reg_model <- glm(Severity_binary ~ Weather_conditions + Driving_experience + Road_allignment + 
                     Type_of_vehicle + Defect_of_vehicle + Area_accident_occured, 
                     data = accidents_clean, 
                     family = binomial)

# Summary of the model to interpret the coefficients
summary(log_reg_model)

# Model diagnostics - Check the significance of predictors
# Look at the coefficients, p-values, and odds ratios
exp(cbind(OR = coef(log_reg_model), confint(log_reg_model)))

# Model Evaluation: Calculate predictions and performance metrics
predictions <- predict(log_reg_model, type = "response")
predicted_classes <- ifelse(predictions > 0.5, 1, 0)

# Confusion matrix and accuracy
conf_matrix <- table(Predicted = predicted_classes, Actual = accidents_clean$Severity_binary)
accuracy <- sum(diag(conf_matrix)) / sum(conf_matrix)
print(paste("Accuracy: ", accuracy))

```

## Answer
The most predictive factors for severe accidents in this dataset are:

1. **Weather Conditions**: Severe accidents are more likely in  "Windy," and "Raining" weather conditions.
2. **Driving Experience**: Drivers with less than 1 year of experience or no license have more severe accident rates.
3. **Road Alignment**: Slopes and more specifically mountainous areas extend risks of fatal accidents significantly..
4. **Vehicle Type**:  Lorries, pick-up vehicles and special vehicles run higher risk when it comes to serious accidents.
5. **Vehicle Defects**:  Some of these vehicle defects can be attributed to a raise slight probability of occurrence of severe accidents. 
The given points leave a possibility of indicating that the probability of heightened accident severity strongly depends on environmental conditions, drivers’ experience, and car types.

# Final Words

Drawing on this road traffic analysis of accidents in Addis Ababa, there are several causes that stand out as determinative of the level of injury occurrence. We noted that weather conditions like ‘Normal’, ‘Windy’, ‘Raining’ have close correlation with severe accidents. Non-experienced/driver with less than one year of license, were more prone to serious accident implying the effect of the driver’s experience. Further, research pointed out that such factors as steep road grades and mountains were some of the factors that boosted accident severity.

The type of vehicle also had it part because lorry, pick-up vehicles, and special types of vehicles from among the vehicles involved contributed to severe accidents. Asehicle problems that compound the risk of fatal injuries, it is important to go for vehicle check-ups.

The study brings out the intensity that environmental factors, the experience of the driver, and the condition of the cars at the time of an accident plays. Preventing very high risk crashes may require policy measures that enhance driver training including first time drivers and vehicle safety inspection. Higher levels of investment in infrastructure could lead to reduction of accidents especially on areas characterized by steep gauges and mountainous regions.

This paper therefore highlights the significance of tailor made safety interventions and may aide in guiding the formulation of road safety polices and engineering designs. More precise data regarding weather conditions, car safety mechanisms, and effects of particular roads’ layout on the severity of an accident can be studied in future researches.

In conclusion it is argued that improvement of the above mentioned factors is crucial in the endeavor of promoting efforts and measures on road safety improvement as well as minimizing the occurrence of serious road accidents in Addis Ababa.


